---
title: "Creating Alluvial Diagrams"
author: "Michał Bojanowski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Alluvial Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, cache=FALSE, include=FALSE}
library(alluvial)
library(dplyr)

knitr::opts_chunk$set(
  fig.show="hold"
)
```

As an example data we will use Titanic data from base. Let's convert it to a data frame:

```{r data}
# Titanic data
tit <- as.data.frame(Titanic, stringsAsFactors = FALSE)
```


# What is alluvial diagram?

```{r}
# Survival status, Sex, and Class
tit %>% group_by(Sex, Class, Survived) %>%
  summarise(n = sum(Freq)) -> tit3d

alluvial(tit3d[,1:3], freq=tit3d$n,
         col = ifelse(tit3d$Survived == "Yes", "orange", "grey")
)
```




# Simple use

By default all alluvia are drawn using gray, mildly transparent colors.




```{r alluvial_defaults}
# Survival status and Class
tit %>% group_by(Class, Survived) %>%
  summarise(n = sum(Freq)) -> tit2d

alluvial(tit2d[,1:2], freq=tit2d$n)

# Survival status, Sex, and Class
tit %>% group_by(Sex, Class, Survived) %>%
  summarise(n = sum(Freq)) -> tit3d

alluvial(tit3d[,1:3], freq=tit3d$n)
```




```{r}
tit3d <- aggregate( Freq ~ Class + Sex + Survived, data=tit, sum)


# 2d
alluvial( tit2d[,1:2], 
          freq=tit2d$Freq, 
          xw=0.0, 
          alpha=0.8,
         gap.width=0.1, col= "steelblue", border="white",
         layer = tit2d$Survived != "Yes" )


alluvial( tit2d[,1:2], freq=tit2d$Freq, 
         hide=tit2d$Freq < 150,
         xw=0.0, alpha=0.8,
         gap.width=0.1, col= "steelblue", border="white",
         layer = tit2d$Survived != "Yes" )

# 3d
tit3d <- aggregate( Freq ~ Class + Sex + Survived, data=tit, sum)

alluvial(tit3d[,1:3], freq=tit3d$Freq, alpha=1, xw=0.2,
         col=ifelse( tit3d$Survived == "No", "red", "gray"),
         layer = tit3d$Sex != "Female",
         border="white")
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
